---
title: "Community Overview"
---

```{r}
#| include: false
source("../params.R")
ghg_emissions <- read_csv("../ghg_emissions.csv") |>
  mutate(entity = case_when(
    entity == "municipal" ~ "Municipal",
    entity == "umass" ~ "UMass", 
    entity == "hampshire_college" ~ "Hampshire College",
    entity == "amherst_college" ~ "Amherst College",
    entity == "community" ~ "Community"
  ),
  entity = factor(entity, levels = c("Hampshire College", "Municipal", "Amherst College", "UMass", "Community")),
  supercategory = factor(supercategory, levels = c("AFOLU", "Waste", "Transportation", "Stationary Energy"))
  )

current_year_emissions <- ghg_emissions |> filter(input_year == current_year) |> pull(total_mtco2e) |> sum(na.rm=TRUE)
baseline_year_emissions <- ghg_emissions |> filter(input_year == baseline_year) |> pull(total_mtco2e) |> sum(na.rm=TRUE)
pct_decrease <- (baseline_year_emissions - current_year_emissions)/baseline_year_emissions*100
```

## Progress to Date

Community-wide emissions have decreased from `r comma(round(baseline_year_emissions,1))` MTCO2e in FY 2016 to `r comma(round(current_year_emissions,1))` in FY 2025, a `r round(pct_decrease,1)`% decrease, falling short of the 25% reduction target for 2025.

```{r}
#| label: fig-community-trend
#| fig-cap: "Plot of community GHG emissions (MTCO2e) by reporting year, including emissions targets set by the Town Council."
community_summary <- ghg_emissions |>
  group_by(input_year) |>
  summarize(total = sum(total_mtco2e, na.rm = TRUE)) 

community_targets <- tibble(
  input_year = c(2016, 2025, 2030, 2050),
  target = c(
  first(community_summary$total),
  first(community_summary$total) * 0.75,
  first(community_summary$total) * 0.5,
  0))

ggplot() +
  geom_point(data = community_summary, aes(input_year, total, color = "Actual Emissions")) +
  geom_line(data = community_summary, aes(input_year, total, color = "Actual Emissions")) +
  geom_point(data = community_targets, aes(input_year, target, color = "Target Emissions")) +
  geom_line(data = community_targets, aes(input_year, target, color = "Target Emissions"), linetype = "dotted") +
  scale_color_manual(values = c("Actual Emissions" = "black", "Target Emissions" = "#8B2635")) +
  labs(y = "MTCO2e") +
  theme(legend.title = element_blank(),
        axis.title.x = element_blank()) + 
  scale_y_continuous(labels = label_comma())
```


```{r}
#| echo: false
#| tbl-cap: Community GHG emissions by year, along with percentage reductions relative to the 2016 baseline.
ghg_emissions |>
  group_by(fiscal_year) |>
  summarize(total = sum(total_mtco2e, na.rm = TRUE)) |>
  mutate(percent_change = percent((total - first(total))/first(total))) |> 
  gt() |>
  cols_label(fiscal_year = "Inventory Year", total = "Community MTCO2e", percent_change = "Percent Reduction") |>
  fmt_number(columns = total, decimals = 0, use_seps = TRUE)
```

## Emissions by Entity

```{r}
p <- ghg_emissions |> 
  group_by(fiscal_year, entity) |> 
  summarize(emissions = sum(total_mtco2e, na.rm=TRUE), .groups = "drop") |> 
  group_by(fiscal_year) |> 
  mutate(pct = emissions/sum(emissions) * 100) |> 
  ggplot(aes(fiscal_year, emissions, fill = entity,
             text = paste0("Entity: ", entity, "<br>",
                           "Emissions: ", comma(round(emissions,1)), "<br>",
                           "Share: ", round(pct, 1), "%"))) +
  geom_col() +
  scale_fill_amherst() +
  labs(y = "MTCO2e",
       fill = NULL) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank()) +
  scale_y_continuous(labels = label_comma())

ggplotly(p, tooltip = "text")
```

```{r}
#| tbl-cap: "Emissions (MTCO2e) per inventory year by entity, along with percent changes from FY 2016 and FY 2022."
ghg_emissions |>
  group_by(input_year, entity) |>
  summarize(emissions = sum(total_mtco2e, na.rm = TRUE), .groups = "drop") |> 
  pivot_wider(names_from = input_year, values_from = emissions) |> 
  mutate(pct_change_2016 = ((`2025` - `2016`) / `2016`),
         pct_change_2022 = ((`2025` - `2022`) / `2022`)) |>
  arrange(desc(`2025`)) |>
  gt() |> 
  cols_label(entity = "Entity", `2016` = "FY 2016", `2022` = "FY 2022", `2025` = "FY 2025",
             pct_change_2016 = "Change Since 2016", pct_change_2022 = "Change Since 2022") |>
  fmt_number(columns = `2016`:`2025`, decimals = 1, use_seps=TRUE) |>
  fmt_percent(columns = pct_change_2016:pct_change_2022, decimals = 1) |>
  data_color(
    columns = c(pct_change_2016, pct_change_2022),
    direction = "column",
    palette = c("#2ca25f", "#ffffff", "#de2d26"), 
    domain = c(-0.7, 0.7)
  )
```

### UMass Amherst

While emissions from UMass Amherst have only modestly decreased since 2016, the university has several major construction projects underway that will lead to significant emissions reductions upon their completion. Some of these projects include:
- North Exchange Energy Center (NEEC) and Thermal Energy Storage (TES). The NEEC has a 640-ton heating/cooling capacity with 70 wells at 800ft depth each.These two projects are at or near completion.

## Emissions by Sector

```{r}
#| label: fig-community-emissions-sector
#| fig-cap: "Emissions (MTCO2e) by sector for each inventory year."
p <- ghg_emissions |> 
  group_by(fiscal_year, supercategory) |> 
  summarize(emissions = sum(total_mtco2e, na.rm=TRUE), .groups = "drop") |> 
  group_by(fiscal_year) |> 
  mutate(pct = emissions/sum(emissions) * 100) |> 
  ggplot(aes(fiscal_year, emissions, fill = supercategory,
             text = paste0("Sector: ", supercategory, "<br>",
                           "Emissions: ", comma(round(emissions,1)), "<br>",
                           "Share: ", round(pct, 1), "%"))) +
  geom_col() +
  scale_fill_amherst() +
  labs(y = "MTCO2e",
       fill = NULL) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank()) +
  scale_y_continuous(labels = label_comma())

ggplotly(p, tooltip = "text")
```

```{r}
#| tbl-cap: "Emissions (MTCO2e) per inventory year by sector, along with percent changes from FY 2016 and FY 2022."
ghg_emissions |>
  group_by(input_year, supercategory) |>
  summarize(emissions = sum(total_mtco2e, na.rm = TRUE), .groups = "drop") |> 
  pivot_wider(names_from = input_year, values_from = emissions) |> 
  mutate(pct_change_2016 = ((`2025` - `2016`) / `2016`),
         pct_change_2022 = ((`2025` - `2022`) / `2022`)) |>
  arrange(desc(`2025`)) |>
  gt() |> 
  cols_label(supercategory = "Sector", `2016` = "FY 2016", `2022` = "FY 2022", `2025` = "FY 2025",
             pct_change_2016 = "Change Since 2016", pct_change_2022 = "Change Since 2022") |>
  fmt_number(columns = `2016`:`2025`, decimals = 1, use_seps=TRUE) |>
  fmt_percent(columns = pct_change_2016:pct_change_2022, decimals = 1) |>
  data_color(
    columns = c(pct_change_2016, pct_change_2022),
    direction = "column",
    palette = c("#2ca25f", "#ffffff", "#de2d26"), 
    domain = c(-0.7, 0.7)
  )
```

