---
title: "Municipal Overview"
---

```{r}
#| include: false
source("../params.R")
ghg_emissions <- read_csv("../ghg_emissions.csv")
municipal_emissions <- ghg_emissions |> filter(entity == "municipal") 
mei_final <- read_csv("../mei_final.csv") |> filter(fiscal_year %in% inventory_years)
```

## Progress to Date

```{r}
#| include: false
baseline <- municipal_emissions |> filter(input_year == baseline_year) |> select(total_mtco2e) |> pull() |> sum()
current <- municipal_emissions |> filter(input_year == current_year) |> select(total_mtco2e) |> pull() |> sum(na.rm=TRUE)
```

The Town of Amherst Municipality was responsible for `r comma(round(baseline, 1))` Metric Tons of CO2 Equivalent emissions in FY 2016. In FY `r current_year`, this number was `r comma(round(current,1))`. This represents a **`r round(100*(baseline-current)/baseline, 1)`% decrease in municipal emissions** relative to the 2016 baseline, which fell short of the Town's 25% emissions reduction goal for 2025. Fortunately, the Town currently has major, net-zero emission, construction and renovation projects underway that will bring municipal emissions much closer to these targets upon their completion. These projects will be discussed later in this report.

```{r}
#| echo: false
#| warning: false
#| label: fig-municipal-trend
#| fig-cap: "Plot of municipal GHG emissions (MTCO2e) by reporting year, and the Town of Amherst's emissions reductions goals with respect to 2016 baseline (25% by 2025, 50% by 2030, 100% by 2050)."
municipal_summary <- municipal_emissions |>
  group_by(input_year) |>
  summarize(total = sum(total_mtco2e, na.rm = TRUE), .groups="drop") 

amherst_targets <- tibble(
  fiscal_year = c(2016, 2025, 2030, 2050),
  target = c(
  first(municipal_summary$total),
  first(municipal_summary$total) * 0.75,
  first(municipal_summary$total) * 0.5,
  0))

ggplot() +
  geom_point(data =municipal_summary, aes(input_year, total, color = "Actual Emissions")) +
  geom_line(data =municipal_summary, aes(input_year, total, color = "Actual Emissions")) +
  geom_point(data = amherst_targets, aes(fiscal_year, target, color = "Target Emissions")) +
  geom_line(data = amherst_targets, aes(fiscal_year, target, color = "Target Emissions"), linetype = "dotted") +
  scale_y_continuous(limits = c(0, 5000)) +
  scale_x_continuous(limits = c(2016, 2050)) +
  scale_color_manual(values = c("Actual Emissions" = "black", "Target Emissions" = "#8B2635")) +
  labs(y = "MTCO2e") +
  theme(legend.title = element_blank(),
        axis.title.x = element_blank())
```



```{r}
#| echo: false
#| tbl-cap: GHG emissions by year, along with percentage reductions relative to the 2016 baseline.
municipal_emissions |>
  group_by(fiscal_year) |>
  summarize(total = sum(total_mtco2e, na.rm = TRUE)) |>
  mutate(percent_change = percent((total - first(total))/first(total))) |> 
  gt() |>
  cols_label(fiscal_year = "Inventory Year", total = "Total MTCO2e", percent_change = "Percent Reduction") |>
  fmt_number(columns = total, decimals = 0, use_seps=TRUE)
```


*Note: Previous inventories reported 4,389 and 3,852 MTCO2e for FY 2016 and FY 2022, respectively. These numbers were updated to reflect the latest data sources and methodologies.*


## Emissions by Sector

Municipal emissions fall into two categories: stationary energy and transportation. Stationary energy accounted for 74% of total emissions in 2025, up from 69% in 2022. 

```{r}
#| echo: false
#| warning: false
#| label: fig-municipal-emissions
#| fig-cap: "Municipal GHG emissions (MTCO2e) by reporting year, excluding transmission and distribution losses"

p <- mei_final |>
  mutate(supercategory = case_when(
    supercategory == "stationary_energy" ~ "Stationary Energy",
    supercategory == "transportation" ~ "Transportation"
  )) |> 
  group_by(fiscal_year_string) |>
  mutate(
    year_total = sum(emission_mtco2e),
    pct = emission_mtco2e/year_total * 100
  ) |>
  group_by(fiscal_year_string, supercategory) |>
  summarize(
   emission_mtco2e = sum(emission_mtco2e),
   pct = sum(pct),
   year_total = first(year_total),
   .groups="drop"
  ) |>
  ggplot(aes(fiscal_year_string, emission_mtco2e, fill = supercategory,
             text = paste0("Sector: ", supercategory, "<br>",
                           "Emissions: ", round(emission_mtco2e, 1), " MTCO2e<br>",
                           "Share: ", round(pct, 1), "%"))) +
  geom_col() +
  scale_fill_amherst() +
  labs(y = "MTCO2e", fill = NULL) +
  theme(axis.title.x = element_blank()) 

ggplotly(p, tooltip = "text")
```

## Emissions by Fuel Type

The composition of fuel usage has changed over time within the municipality. Most notably, electricity has seen an increased share of total emissions, while the relative shares of gasoline and diesel have decreased. This is consistent with the Town's move toward more renewable energy sources.

```{r}
#| label: fig-fuel-mix
#| fig-cap: "Share of greenhouse gas emissions by fuel source, excluding transmission and distribution losses."

mei_interactive <- mei_final |>
  group_by(fiscal_year_string, account_fuel) |>
  summarize(total_emissions = sum(emission_mtco2e), .groups = "drop") |>
  group_by(fiscal_year_string) |>
  mutate(
    percentage = total_emissions / sum(total_emissions) * 100,
    tooltip_text = paste0(
      "Fuel: ", account_fuel, "\n",
      "Emissions: ", round(total_emissions, 2), " MTCO2e\n",
      "Share: ", round(percentage, 1), "%"
    )
  )
p <- ggplot(mei_interactive, aes(x = fiscal_year_string, y = total_emissions, text = tooltip_text)) +
  geom_col(position = "fill", aes(fill = account_fuel)) +
  scale_fill_amherst() +
  labs(fill = NULL) +
  theme(legend.title = element_blank(),
        axis.title = element_blank())
ggplotly(p, tooltip = "text")
```

```{r}
#| tbl-cap: "Fuel consumption in original units, with percent change relative to 2016 and 2022 inventories."
mei_final |>
  group_by(fiscal_year, account_fuel, updated_units) |>
  summarize(total_consumption = sum(use_updated_units), .groups = "drop") |>
    mutate(sector = case_when(
    account_fuel %in% c("Gasoline", "Diesel") ~ "Transportation",
    TRUE ~ "Buildings & Stationary Infrastructure"
  )) |>
  pivot_wider(names_from = fiscal_year, values_from = total_consumption) |>
  rename(current = !!sym(as.character(current_year))) |>
  mutate(pct_change_2016 = ((current - `2016`) / `2016`),
         pct_change_2022 = ((current - `2022`) / `2022`),
         updated_units = case_when(
           updated_units == "MWh" ~ updated_units,
           .default = str_to_title(updated_units)
         )) |>
  gt(groupname_col = "sector") |>
  tab_options(
    row_group.font.weight = "bold", 
    row_group.font.size = px(15),       
    row_group.background.color = "#f9f9f9"
  ) |>
  cols_label(account_fuel = "Fuel Type", updated_units = "Units", 
             `2016` = "FY 2016", `2022` = "FY 2022", current = paste0("FY ", current_year),
             pct_change_2016 = "Change Since 2016", pct_change_2022 = "Change Since 2022") |>
  fmt_number(columns = c(`2016`, `2022`, current), decimals = 1, use_seps=TRUE) |>
  fmt_percent(columns = pct_change_2016:pct_change_2022, decimals = 1) |>
  data_color(
    columns = c(pct_change_2016, pct_change_2022),
    direction = "column",
    palette = c("#2ca25f", "#ffffff", "#de2d26"), 
    domain = c(-0.4, 0.4)
  )
```

