---
title: "Municipal Stationary Energy"
---

```{r}
#| include: false
source("../params.R")
library(tidyverse)
mei_final <- read_csv("../mei_final.csv") 
mei_stationary <- mei_final %>% filter(fiscal_year %in% c(2016, 2022, 2025), supercategory == "stationary_energy")
```

## Broad Trends in Stationary Emissions

Stationary emissions accounted for 3,195 MTCO2e in 2016, and 2,704 MTCO2e in 2025, representing a **15.4% reduction**. Notably, despite an overall reduction in municipal emissions since 2022, stationary energy emissions were slightly higher in 2025 than they were in 2022, an increase of 78 MTCO2e. 

```{r}
#| echo: false
#| label: fig-stationary-emission-trends
#| fig-cap: "Municipal stationary emissions (MTCO2e) trends by fiscal year."


mei_final %>% 
  filter(fiscal_year >= 2016, supercategory == "stationary_energy") %>%
  group_by(fiscal_year_string, inventory_year) %>%
  summarize(total_emissions = sum(emission_mtco2e)) %>%
  ggplot(aes(fiscal_year_string, total_emissions, group = 1)) +
  geom_line() +
  geom_point(
    data = ~ filter(.x, inventory_year == "0"),
    color = "black",
    size = 2
  ) +
  geom_point(
    data = ~ filter(.x, inventory_year == "1"),
    color = "#C9A040",
    size = 5,
    shape = 18
  ) +
  labs(y = "MTCO2e") +
  theme(
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  scale_y_continuous(limits = c(2250, 3500))
```


```{r}
#| echo: false
#| tbl-cap: Stationary GHG emissions by year, along with percentage reductions relative to the 2016 baseline.
mei_stationary %>%
  group_by(fiscal_year) %>%
  summarize(total = sum(emission_mtco2e, na.rm = TRUE)) %>%
  mutate(percent_change = percent((total - first(total))/first(total))) %>% 
  gt() %>%
  cols_label(fiscal_year = "Inventory Year", total = "Stationary MTCO2e", percent_change = "Percent Reduction") %>%
  fmt_number(columns = total, decimals = 0, use_seps=TRUE)
```

### Emissions by Fuel Type

```{r}
#| label: fig-stationary-fuel-type
#| fig-cap: "Stationary emissions by year by fuel type"

p <- mei_final %>%
  filter(fiscal_year >= 2016, supercategory == "stationary_energy") %>%
  mutate(account_fuel = factor(account_fuel, levels = c("Propane", "Oil", "Gas", "Electric")),
         inventory_year = as.character(inventory_year)) %>%
  group_by(fiscal_year_string, inventory_year, account_fuel) %>%
  summarize(total_emissions = sum(emission_mtco2e), .groups = "drop") %>%
  group_by(fiscal_year_string) %>%
  mutate(pct = total_emissions/sum(total_emissions)*100,
         tooltip = paste0("Fuel: ", account_fuel, "<br>",
                          "Emissions: ", total_emissions, "<br>",
                          "Share: ", round(pct,1), "%<br>")) %>%
  ggplot(aes(fiscal_year_string, total_emissions, 
             fill = account_fuel, text = tooltip,
             alpha = inventory_year)) +
  geom_col() + 
  scale_alpha_manual(values = c("1" = 1, "0" = 0.7)) +
  scale_fill_amherst() +
  labs(y = "MTCO2e") +
  theme(legend.title = element_blank(),
        axis.title.x = element_blank())

plt <- ggplotly(p, tooltip = "text") %>%
  layout(legend = list(title=list(text="")))

plt$x$data[[5]]$name <- "Propane"
plt$x$data[[6]]$name <- "Oil"
plt$x$data[[7]]$name <- "Methane Gas"
plt$x$data[[8]]$name <- "Electricity"

for (i in 1:4)  plt$x$data[[i]]$showlegend <- FALSE
for (i in 5:8)  plt$x$data[[i]]$showlegend <- TRUE

plt
```


### Emissions by Facility Category

Buildings are responsible for the largest share of municipal emissions, accounting for 59.5% of stationary emissions for FY 2025. The next largest share is accounted for by water and sewer facilities, which amounts to 36.6% in FY 2025. The remaining emissions are attributed to open spaces and street/traffic lights, which account for the remaining 3.9%.


```{r}
#| echo: false
#| warning: false
#| label: fig-building-category
#| fig-cap: "Municipal GHG emissions (MTCO2e) by building category"

p <- mei_stationary %>%
  group_by(fiscal_year_string) %>%
  mutate(year_total = sum(emission_mtco2e),
    pct = emission_mtco2e/year_total * 100) %>%
  group_by(fiscal_year_string, building_category) %>%
  summarize(emission_mtco2e = sum(emission_mtco2e),
   pct = sum(pct),
   year_total = first(year_total),
   .groups="drop") %>% 
  mutate(building_category = factor(building_category, levels = c("Open Space",  "Street/Traffic Lights", "Water/Sewer", "Building"))) %>%
  ggplot(aes(fiscal_year_string, emission_mtco2e, fill = building_category,
             text = paste0("Category: ", building_category, "<br>",
                           "Emissions: ", round(emission_mtco2e, 1), " MTCO2e<br>",
                           "Share: ", round(pct, 1), "%"))) +
  geom_col() +
  scale_fill_amherst() +
  labs(y = "MTCO2e", fill = NULL) +
  theme(axis.title.x = element_blank())

ggplotly(p, tooltip = "text")
```


## Electricity

Electricity consumption increased significantly from 5674 megawatt hours (MWh) in FY 2022 to 6242 MWh in FY 2025, though this still represents a decrease from the FY 2016 consumption of 6398 MWh. Electricity consumption has increased steadily since the sharp decline in FY 2020 and FY 2021 due to the COVID-19 pandemic, and current levels in FY 2025 are much more typical of pre-pandemic usage. 

```{r}
#| echo: false
#| label: fig-electricity-trends
#| fig-cap: "Municipal electricity usage (MWh) trends by fiscal year."


mei_final %>% 
  filter(account_fuel == "Electric", fiscal_year >= 2016) %>%
  group_by(fiscal_year_string, inventory_year, account_fuel, updated_units) %>%
  summarize(total_usage = sum(use_updated_units)) %>%
  ggplot(aes(fiscal_year_string, total_usage, group = 1)) +
  geom_line() +
  geom_point(data = ~ filter(.x, inventory_year == "0"),
    color = "black",
    size = 2) +
  geom_point(data = ~ filter(.x, inventory_year == "1"),
    color = "#C9A040",
    size = 5,
    shape = 18) +
  labs(y = "Usage (MWh)") +
  theme(legend.position = "none",
    axis.title.x = element_blank()) +
  scale_y_continuous(limits = c(4500, 7000))
```

### Electricity Top Consumers


```{r}
#| label: fig-electric-consumers
#| fig-cap: "Highest electricity-consuming facilities (MWh) in 2025."
mei_final %>%
  filter(account_fuel == "Electric", fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(electricity_usage = sum(use_updated_units), .groups="drop") %>%
  arrange(desc(electricity_usage)) %>%
  slice_max(electricity_usage, n = 10) %>%
  ggplot(aes(electricity_usage, forcats::fct_reorder(building, electricity_usage))) +
  geom_col(fill = "#C9A040") + 
  labs(x = "Electricity Usage (MWh)") +
  theme(axis.title.y = element_blank())
```

```{r}
#| label: tbl-electricity-consumers
#| tbl-cap: "Highest electricity-consuming (MWh) facilities in 2025, with their associated emissions (MTCO2e)."

mei_final %>%
  filter(account_fuel == "Electric", fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(electricity_usage = sum(use_updated_units), 
            emissions_mtco2e = sum(emission_mtco2e),
            .groups="drop") %>%
  slice_max(electricity_usage, n=10) %>%
  gt() %>%
  cols_label(building = "Facility", electricity_usage = "Electricity Usage (MWh)", emissions_mtco2e = "Emissions (MTCO2e)") %>%
  fmt_number(columns = electricity_usage:emissions_mtco2e, decimals = 1, use_seps = TRUE)
```


## Methane Gas

```{r}
#| echo: false
#| label: fig-gas-trends
#| fig-cap: "Municipal methane gas usage (therms) by fiscal year."


mei_final %>% 
  filter(account_fuel == "Gas", fiscal_year >= 2016) %>%
  group_by(fiscal_year_string, inventory_year, account_fuel, updated_units) %>%
  summarize(total_usage = sum(use_updated_units)) %>%
  ggplot(aes(fiscal_year_string, total_usage, group = 1)) +
  geom_line() +
  geom_point(data = ~ filter(.x, inventory_year == "0"),
    color = "black",
    size = 2) +
  geom_point(data = ~ filter(.x, inventory_year == "1"),
    color = "#C9A040",
    size = 5,
    shape = 18) +
  labs(y = "Usage (therms)") +
  theme(legend.position = "none",
    axis.title.x = element_blank()) +
  scale_y_continuous(limits = c(90000, 130000))
```
### Methane Gas Top Consumers

```{r}
#| label: fig-gas-consumers
#| fig-cap: "Highest methane gas-consuming facilities (therms) in 2025."
mei_final %>%
  filter(account_fuel == "Gas", fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(usage = sum(use_updated_units), .groups="drop") %>%
  arrange(desc(usage)) %>%
  slice_max(usage, n = 10) %>%
  ggplot(aes(usage, forcats::fct_reorder(building, usage))) +
  geom_col(fill = "#5A7A5E") + 
  labs(x = "Methane Gas Usage (therms)") +
  theme(axis.title.y = element_blank())
```

```{r}
#| label: tbl-gas-consumers
#| tbl-cap: "Highest methane gas-consuming (therms) facilities in 2025, with their associated emissions (MTCO2e)."

mei_final %>%
  filter(account_fuel == "Gas", fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(usage = sum(use_updated_units), 
            emissions_mtco2e = sum(emission_mtco2e),
            .groups="drop") %>%
  slice_max(usage, n=10) %>%
  gt() %>%
  cols_label(building = "Facility", usage = "Methane Gas Usage (therms)", emissions_mtco2e = "Emissions (MTCO2e)") %>%
  fmt_number(columns = usage:emissions_mtco2e, decimals = 1, use_seps = TRUE)
```


## Oil

```{r}
#| echo: false
#| label: fig-oil-trends
#| fig-cap: "Municipal oil usage (gallons) by fiscal year."


mei_final %>% 
  filter(account_fuel == "Oil", fiscal_year >= 2016) %>%
  group_by(fiscal_year_string, inventory_year, account_fuel, updated_units) %>%
  summarize(total_usage = sum(use_updated_units)) %>%
  ggplot(aes(fiscal_year_string, total_usage, group = 1)) +
  geom_line() +
  geom_point(data = ~ filter(.x, inventory_year == "0"),
    color = "black",
    size = 2) +
  geom_point(data = ~ filter(.x, inventory_year == "1"),
    color = "#C9A040",
    size = 5,
    shape = 18) +
  labs(y = "Usage (gallons)") +
  theme(legend.position = "none",
    axis.title.x = element_blank()) +
  scale_y_continuous(limits = c(50000, 100000))
```


```{r}
#| label: fig-oil-consumers
#| fig-cap: "Highest oil-consuming facilities (gallons) in 2025."
mei_final %>%
  filter(account_fuel == "Oil", fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(usage = sum(use_updated_units), .groups="drop") %>%
  arrange(desc(usage)) %>%
  slice_max(usage, n = 10) %>%
  ggplot(aes(usage, forcats::fct_reorder(building, usage))) +
  geom_col(fill = "#2E7D8C") + 
  labs(x = "Oil Usage (gallons)") +
  theme(axis.title.y = element_blank())
```

```{r}
#| label: tbl-oil-consumers
#| tbl-cap: "Highest oil-consuming (gallons) facilities in 2025, with their associated emissions (MTCO2e)."

mei_final %>%
  filter(account_fuel == "Oil", fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(usage = sum(use_updated_units), 
            emissions_mtco2e = sum(emission_mtco2e),
            .groups="drop") %>%
  slice_max(usage, n=10) %>%
  gt() %>%
  cols_label(building = "Facility", usage = "Oil Usage (gallons)", emissions_mtco2e = "Emissions (MTCO2e)") %>%
  fmt_number(columns = usage:emissions_mtco2e, decimals = 1, use_seps = TRUE)
```


## Highest-Emitting Facilities

```{r}
top_emitters <- mei_stationary %>%
  filter(fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(total_emissions = sum(emission_mtco2e)) %>%
  arrange(desc(total_emissions)) %>%
  head(10) %>%
  pull(building)

top_building_emissions <- mei_stationary %>%
  filter(building %in% top_emitters, fiscal_year == 2025) %>%
  summarize(total_emissions = sum(emission_mtco2e)) %>%
  pull(total_emissions)
 
emissions_2025 <- mei_final %>% filter(fiscal_year == 2025) %>% pull(emission_mtco2e) %>% sum()
stationary_emissions_2025 <-  mei_final %>% 
  filter(fiscal_year == 2025, supercategory == "stationary_energy") %>% 
  pull(emission_mtco2e) %>% 
  sum()
school_emissions_2025 <- mei_final %>% 
  filter(fiscal_year == 2025, supercategory == "stationary_energy", building %in% c("Wildwood ES", "Fort River ES")) %>% 
  pull(emission_mtco2e) %>% 
  sum()

```


- The 10 municipal facilities with the highest emissions account for `r round(top_building_emissions, 1)` MTCO2e in 2025, which accounts for `r round(top_building_emissions/emissions_2025 * 100)`% of total municipal emissions.
- The new Amethyst Brook Elementary School is scheduled to be completed by the end of 2026, and will be a **net zero** facility. This building will replace both the Wildwood and Fort River elementary schools, which together account for `r round(school_emissions_2025, 1)` MTCO2e in FY 2025.
- The Jones Library renovation project is expected to be completed in early 2027. This is another **net zero** facility, which accounts for 119 MTCO2e in FY 2025, prior to renovations.
- C

Wildwood, Fort River, and Crocker Farm Elementary Schools combine for `r round(school_emissions_2025, 1)` MTCO2e emissions in 2025, which amounts to `r round(school_emissions_2025/stationary_emissions_2025*100, 1)`% of municipal stationary emissions in 2025, or **`r round(school_emissions_2025/emissions_2025*100, 1)`% of total municipal emissions in 2025.**

Wildwood, Fort River, and Crocker Farm Elementary Schools combine for `r round(school_emissions_2025, 1)` MTCO2e emissions in 2025, which amounts to `r round(school_emissions_2025/stationary_emissions_2025*100, 1)`% of municipal stationary emissions in 2025, or **`r round(school_emissions_2025/emissions_2025*100, 1)`% of total municipal emissions in 2025.** With the completion of construction of the new elementary school !!add name and details!!, which has been designed to be much more energy efficient, we expect these numbers to fall (by how much? can we estimate this?). **Also talk about the Centennial building, if that would account for the increase in emissions for Pump Stations**

```{r}
#| tbl-cap: Emissions (MTCO2e) data for highest-emissions facilities, with percent change in emissions from both 2016 and 2022 inventories.
top_emitters <- mei_stationary %>%
  filter(fiscal_year == 2025) %>%
  group_by(building) %>%
  summarize(total_emissions = sum(emission_mtco2e)) %>%
  arrange(desc(total_emissions)) %>%
  head(10) %>%
  pull(building)

mei_stationary %>% 
  filter(building %in% top_emitters) %>%
  group_by(fiscal_year, building, building_subcategory) %>% 
  summarize(emissions = sum(emission_mtco2e), .groups = "drop") %>%
  pivot_wider(names_from = fiscal_year, values_from = emissions) %>%
  mutate(pct_change_2016 = ((`2025` - `2016`) / `2016`),
         pct_change_2022 = ((`2025` - `2022`) / `2022`)) %>%
  arrange(desc(`2025`)) %>%
  gt() %>%
  cols_label(building = "Facility", building_subcategory = "Facility Subcategory", 
             `2016` = "FY 2016", `2022` = "FY 2022", `2025` = "FY 2025",
             pct_change_2016 = "Change Since 2016", pct_change_2022 = "Change Since 2022") %>%
  fmt_number(columns = `2016`:`2025`, decimals = 1, use_seps=TRUE) %>%
  fmt_percent(columns = pct_change_2016:pct_change_2022, decimals = 1) %>%
  data_color(
    columns = c(pct_change_2016, pct_change_2022),
    direction = "column",
    palette = c("#2ca25f", "#ffffff", "#de2d26"), 
    domain = c(-0.7, 0.7)
  )
```


