---
title: "Municipal Transportation"
---

```{r}
#| include: false
source("../params.R")
mei_final <- read_csv("../mei_final.csv") 
mei_transportation<- mei_final |> filter(supercategory == "transportation")

mei_transportation <- mei_transportation |>
  mutate(
    fleet_group = case_when(
      is.na(usage_reading_type) & department == "Police Department" ~ "Police",
      is.na(usage_reading_type) & department == "Public Works" ~ "Public Works",
      str_to_lower(usage_reading_type) %in% c("fire") ~ "Fire",
      str_to_lower(usage_reading_type) %in% c("police") ~ "Police",
      str_to_lower(usage_reading_type) %in% c("cons") ~ "Conservation",
      str_to_lower(usage_reading_type) %in% c("maint") ~ "Maintenance",
      str_to_lower(usage_reading_type) %in% c("school", "schools") ~ "Schools",
      str_to_lower(usage_reading_type) %in% c("dpw") ~ "Public Works",
      str_to_lower(usage_reading_type) %in% c("library", "lib") ~ "Library",
      str_to_lower(usage_reading_type) %in% c("coa", "sen ctr", "sr ctr") ~ "Senior Center",
      str_to_lower(usage_reading_type) %in% c("rec") ~ "Recreation",
      str_to_lower(usage_reading_type) %in% c("cherry hill", "ch hill") ~ "Cherry Hill Golf Course",
      str_to_lower(usage_reading_type) %in% c("it") ~ "Information Technology",
      str_to_lower(usage_reading_type) %in% c("lsse") ~ "LSSE",
      str_to_lower(usage_reading_type) %in% c("cress") ~ "CRESS",
      TRUE ~ usage_reading_type
    )
  )
```

## Broad Trends in Transportation Emissions

Municipal emissions related to transportation decreased from 1,230 MTCO2e in FY 2016 to 955 MTCO2e in FY 2025, representing a **22.3% decrease** over that period.

```{r}
#| echo: false
#| label: fig-transportation-emission-trends
#| fig-cap: "Municipal transportation emissions (MTCO2e) trends by fiscal year."


mei_final |> 
  filter(fiscal_year >= 2016, supercategory == "transportation") |>
  group_by(fiscal_year_string, inventory_year) |>
  summarize(total_emissions = sum(emission_mtco2e)) |>
  ggplot(aes(fiscal_year_string, total_emissions, group = 1)) +
  geom_line() +
  geom_point(
    data = ~ filter(.x, inventory_year == "0"),
    color = "black",
    size = 2
  ) +
  geom_point(
    data = ~ filter(.x, inventory_year == "1"),
    color = "#C9A040",
    size = 5,
    shape = 18
  ) +
  labs(y = "MTCO2e") +
  theme(
    legend.position = "none",
    axis.title.x = element_blank()
  ) + 
  scale_y_continuous(limits = c(800, 1400))
```


```{r}
#| echo: false
#| tbl-cap: Transportation GHG emissions by year, along with percentage reductions relative to the 2016 baseline.
mei_transportation |>
  filter(inventory_year == 1) |>
  group_by(fiscal_year) |>
  summarize(total = sum(emission_mtco2e, na.rm = TRUE)) |>
  mutate(percent_change = percent((total - first(total))/first(total))) |> 
  gt() |>
  cols_label(fiscal_year = "Inventory Year", total = "Transportation MTCO2e", percent_change = "Percent Reduction") |>
  fmt_number(columns = total, decimals = 0, use_seps=TRUE)
```


```{r}
#| label: fig-transportation-fuel
#| fig-cap: "Municipal transportation emissions per year by fuel type."

p <- mei_final |>
  filter(fiscal_year >= 2016, supercategory == "transportation") |>
  mutate(account_fuel = factor(account_fuel, levels = c("Diesel", "Gasoline")),
         inventory_year = as.character(inventory_year)) |>
  group_by(fiscal_year_string, inventory_year, account_fuel) |>
  summarize(total_emissions = sum(emission_mtco2e), .groups = "drop") |>
  group_by(fiscal_year_string) |>
  mutate(pct = total_emissions/sum(total_emissions)*100,
         tooltip = paste0("Fuel: ", account_fuel, "<br>",
                          "Emissions: ", total_emissions, "<br>",
                          "Share: ", round(pct,1), "%<br>")) |>
  ggplot(aes(fiscal_year_string, total_emissions, 
             fill = account_fuel, text = tooltip,
             alpha = inventory_year)) +
  geom_col() + 
  scale_alpha_manual(values = c("1" = 1, "0" = 0.7)) +
  scale_fill_amherst() +
  labs(y = "MTCO2e") +
  theme(legend.title = element_blank(),
        axis.title.x = element_blank())

plt <- ggplotly(p, tooltip = "text") |>
  layout(legend = list(title=list(text="")))

plt$x$data[[3]]$name <- "Diesel"
plt$x$data[[4]]$name <- "Gasoline"

for (i in 1:2)  plt$x$data[[i]]$showlegend <- FALSE
for (i in 3:4)  plt$x$data[[i]]$showlegend <- TRUE

plt
```


## Transportation Emissions by Fleet Group

96% of transportation-related emissions can be accounted for by the four highest-consumption fleet categories: Public Works, Fire Department, Police Department, and Schools. With the exception of Fire Department vehicles, each of these fleet groups have decreased their emissions since FY 2016. 

```{r}
#| label: fig-fleet-group-trends
#| fig-cap: "Municipal emissions (MTCO2e) by fleet group over time."
mei_transportation |> 
  filter(fleet_group %in% c("Public Works", "Fire", "Police", "Schools"), fiscal_year >= 2016) |> 
  group_by(fiscal_year_string, fleet_group) |> 
  summarize(emissions = sum(emission_mtco2e), .groups="drop") |> 
  mutate(fleet_group = fct_reorder2(fleet_group, fiscal_year_string, emissions)) |>
  ggplot(aes(fiscal_year_string, emissions, color = fleet_group, group = fleet_group)) + 
  geom_point() + 
  geom_line() + 
  scale_color_amherst() +
  labs(y = "MTCO2e") +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 0))
```


```{r}
mei_transportation |> 
  filter(fiscal_year %in% c(2016, 2022, 2025)) |>
  group_by(fiscal_year, fleet_group) |> 
  summarize(emissions = sum(emission_mtco2e), .groups = "drop") |>
  pivot_wider(names_from = fiscal_year, values_from = emissions) |>
  mutate(pct_change_2016 = ((`2025` - `2016`) / `2016`),
         pct_change_2022 = ((`2025` - `2022`) / `2022`)) |>
  arrange(desc(`2025`)) |>
  slice_max(`2025`, n=6) |>
  gt() |>
  cols_label(fleet_group = "Fleet Group", 
             `2016` = "FY 2016", `2022` = "FY 2022", `2025` = "FY 2025",
             pct_change_2016 = "Change Since 2016", pct_change_2022 = "Change Since 2022") |>
  fmt_number(columns = `2016`:`2025`, decimals = 1, use_seps=TRUE) |>
  fmt_percent(columns = pct_change_2016:pct_change_2022, decimals = 1) |>
  data_color(columns = c(pct_change_2016, pct_change_2022),
    direction = "column",
    palette = c("#2ca25f", "#ffffff", "#de2d26"), 
    domain = c(-0.7, 0.7))
```

